---
title: "Coding task 21.01.2021: Comapring distributions"
output:
  html_document: 
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Example data & libraries
```{r}
# install.packages("rattle.data")
# install.packages("ggridges")
library(rattle.data)
library(ggplot2)
library(ggpubr)

wine = wine

```

# Different ways to visualise distributions

## Plot the distribution of the wine color for the three different wine types
- Try to come up with different way to plot this
- Which is you favorite and why?

\newpage
## Here are some more, try them out

### geom_density(), ggridges::geom_density_ridges(), geom_histogram()
- Try out different bins for the histogram
- What is the difference of a histogram to a density plot?

```{r}
ggplot(wine, aes(x =Color, color = Type))+
  geom_density()

ggplot(wine, aes(x =Color))+
  geom_histogram(aes( fill = Type))

ggplot(wine, aes(x =Color))+
  geom_histogram(aes( fill = Type), position = "dodge")

ggplot(wine, aes(x =Color))+
  geom_histogram(aes(fill = Type), bins = 50)

ggplot(wine, aes(x =Color))+
  geom_histogram(aes(fill = Type), bins = 2)

ggplot(wine, aes(x =Color, y = Type))+
  ggridges::geom_density_ridges()

```


### geom_boxplot(), geom_jitter(), geom_violin()
- How do these plots relate to a classical density plot?
- What information do you lose in which plot?
- You can also combine these three or two of them

```{r}

ggplot(wine, aes(x =Type, y = Color))+
  geom_boxplot()

ggplot(wine, aes(x =Type, y = Color))+
  geom_jitter()

ggplot(wine, aes(x =Type, y = Color))+
  geom_boxplot()+
  geom_jitter()

ggplot(wine, aes(x =Type, y = Color))+
  geom_violin()

ggplot(wine, aes(x =Type, y = Color))+
  geom_violin()+
  geom_boxplot()

ggplot(wine, aes(x =Type, y = Color))+
  geom_violin()+
  geom_jitter()

```


### stat_ecdf()
- What does this plot show you?
```{r}
ggplot(wine, aes(x =Color, fill = Type))+
  stat_ecdf()


```

# Statistical tests to compare distributions (of continous values)
## What is a p-value? How would you explain it?
## compare_means(), stat_compare_means()

- Compare_means() statisticaly compares two values
- stat_compare_means() can be added directly to a ggplot with plot + stat_compare_means() and adds the p-values to the plot

### make a box or violin plot from above but only comparing wines of type 1 and 2 and test both functions

- Which statistical test is used by default? 
- How do the p-values change, if you use method = "t-test" 
- When would you use "t-test" as method instead? 
- Use ggqqplot() and the shapiro.test() to decide if you could use a t-test here

```{r}

compare_means(Color ~ Type, data=wine[wine$Type !=3,] )

# ! attention 
# compare_means(Type ~ Color, data=wine[wine$Type !=3,] ) does not work x~y, x must be numeric y is the grouping

ggplot(wine[wine$Type !=2,], aes(x =Type, y = Color))+
  geom_boxplot()+
  stat_compare_means()+
  ggtitle( "The standard test for two groups is a Wilcoxon test")

ggplot(wine[wine$Type !=2,], aes(x =Type, y = Color))+
  geom_boxplot()+
  stat_compare_means(method = "t.test")+
  ggtitle("T-test sometimes give other p-values as they assume normal distributions", subtitle = "We need to test for normality if we want to use a t-test")


##########
# test for normality
#########
shapiro.test(wine[wine$Type==1,]$Color)
shapiro.test(wine[wine$Type==2,]$Color)
shapiro.test(wine[wine$Type==3,]$Color)
shapiro.test(wine$Color)
# Null hypothesis: the vector is normaly distributed, p-value: gives you the probablility that the vector was normally distributed
# The smaller the p-value the smaller the probability of a normal distribution
# here all vectors are likely non-normal distributed

ggqqplot(wine[wine$Type==1,]$Color)+
  ggtitle("this plot shows a theoretical normal distribution and our values in it")
ggqqplot(wine[wine$Type==2,]$Color)
ggqqplot(wine[wine$Type==3,]$Color)

```




## Pairwise comparisons for three groups
- If you now make a box/violin plot with all three types of wine, what is the default test? Why is the default different now?
- Instead you can do three pairwise comparisons:
- make a list of your comparisons: my_comparisons <- list( c("3", "1"), c("1", "2"), c("3", "2") )
- use stat_compare_means(comparisons = my_comparisons)

```{r}
ggplot(wine, aes(x =Type, y = Color))+
  geom_boxplot()+
  stat_compare_means()+
  ggtitle("If we compare more then two groups the default test is a Kruksal-Walis test, \n if normal distributed we could use a anova test")

my_comparisons <- list( c("3", "1"), c("1", "2"), c("3", "2") )

ggplot(wine, aes(x =Type, y = Color))+
  geom_boxplot()+
  stat_compare_means(comparisons = my_comparisons)+
  ggtitle("We can compare pairs of to groups instead again with Wilcoxon ")

ggplot(wine, aes(x =Type, y = Color))+
  geom_boxplot()+
  stat_compare_means(comparisons = my_comparisons, method = "t.test")+
  ggtitle("Or with the t.test if we have a normal distribution \n (we saw before that thats not the case here)")


```





