---
title: "Coding task 12.03.21 - Looping over lists"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Lists in R
- in lists you can sum up all other types of data in R
- e.g a list of data frames, a list of vectors, a list of lists
- the list items do not have to have the same dimensions
- very usefull if you have two similar data sets, that you want to treat the same way (e.g two conditions/samples)
```{r}
library(purrr)
library(dplyr)

# example data
data("beavers")
data("airquality")
data("mdeaths")

# we have data from two beavers
# beaver1
# beaver2

# and put both together in a list
beavers <- list(beaver1, beaver2)

# you can give each item within a list a name
names(beavers)<- c("Beaver1", "Beaver2")

# you can access list items via [[]]
head(beavers[[1]])

# or via the name
head(beavers$Beaver2)

```

**Task 1:** Build a List containing the beaver list and the airquality data set and maybe a random word of you choice.

```{r}
l <- list(beavers, airquality, "Hello List")
names(l)<- c("A", "B", "C")

```



# Loops in R
If you want to loop over lists it is of course helpfull is they have a similar layout.
Here are three ways to loop over a list:

## For Loops
The easiest loop in R is s for loop. (However, for loops are often the most coding intensive and slowest type of loops.) Here is an example on how to use a for loop on a data set
```{r}
# initialise a variable, where you want to save the results from the loop
mean_temp <- vector()

for(i in 1:length(beavers)) # here you specify the loop variable and how often the loop is run (from 1 to the length of the list = 2)
  {
  mean_temp[i] <- mean(beavers[[i]]$temp)
}


# note: of course with two you could still calcualte it seperatly, but you might have 10 beavers :)


```

## Apply, sapply, lapply
These are base R functionalities that are faster and less text intensive then for loops. Their differ in thei output type:
- lapply return a list
- sapply returns a vector 
- apply returns a dataframe
```{r}

# lapply, sapply
# takes as input a list and a function of what to do with each list element, here the function variable x stands for the first listentry in the first round, the second in the second and so on
mean_temp2_list <- lapply(beavers, function(x) mean(x$temp))
mean_temp2_vector <- sapply(beavers, function(x) mean(x$temp))

# apply cannot be used directly on the list but needs a data frame
# but you can loop over all columns at the same time
# you can specify 1 for apply this function rowwise or 2 for column wise
all_means_beaver1 <- apply(beaver1, 2, mean)

# you can combine laply an apply to get loop over all columns of all list menbers
all_mean_both_beavers <- lapply(beavers, apply,2, mean)

```

## purrr::map
This is a tidyverse alternative to loop over lists.
```{r eval=F}
library(purrr)

# map marks the function by the ~ sign and within the function the variable is always caled .x
map(beavers, ~mean(.x$temp)) 
beavers %>% map( ~mean(.x$temp)) 

# you can again change what kind of output you want to get
# a numeric vector
map_dbl(beavers, ~mean(.x$temp))

# a character vector
map_chr(beavers, ~mean(.x$temp))

# map allows you to use dplyr within each loop
library(dplyr) ## You made a nice task on dplyr if you are not familiar with it, it might be worth looking at
map(beavers, ~mutate(.x, hour = substring(as.character(time),1,1)))

# and to use pipes within the loop
map(beavers, ~mutate(.x, hour = substring(as.character(time),1,1)) %>% 
     arrange(hour))

```

## Try to solve these tasks with all three methods

**Task 2** Calucalte the sd of the 2 columns of the 2 dataframes.
```{r}
###################
# for loop
#################
sd_temp <- vector()

for(i in 1:length(beavers)) 
  {
  sd_temp[i] <- sd(beavers[[i]]$temp)
}

sd_temp

################
# sapply
################
sd_temp2 <- lapply(beavers, function(x) sd(x$temp))
sd_temp2

#################
# map
#################
sd_temp3 <- map_dbl(beavers, ~sd(.x$temp))
sd_temp3 

```

**Task 2b** Add up the time and the temperature.
```{r}
###################
# for loop
#################
add <- vector()

for(i in 1:length(beavers)) 
  {
  add[i] <- beavers[[i]]$temp + beavers[[i]]$time
}

add

################
# sapply
################
add2 <- lapply(beavers, function(x) x$temp = x$time)
add2

#################
# map
#################
add3 <- map(beavers, ~ (.x$temp + .x$time))
add3

add4 <- map(beavers, ~.x %>% mutate(., add = temp + time))
add4 %>% map(~head(.x))




```


**Task 3** Calcluate the rowwise mean (athough its meaningless here).

```{r}
##############
# for loops
############
rowmean <- list()

for(i in 1:length(beavers))  {
      rowmean[[i]] <- list()
  for(j in 1:nrow(beavers[[i]])){
  rowmean[[i]][[j]] <- mean(as.numeric(beavers[[i]][j,]))
}}

rowmean %>% map(~head(.x))



rowmean2 <- list()
for(i in 1:length(beavers)) {
  rowmean2[[i]] <- rowMeans(beavers[[i]])
}

rowmean2 %>% map(~head(.x))

##############
# apply
############
rowmean3 <-lapply(beavers, apply, 1, mean)
rowmean3 %>% map(~head(.x))



################
# map
##############

rowmean4 <- beavers %>% map(~mutate(.x, means = rowMeans(.)))
rowmean4 %>% map(~head(.x))

# and a somewhat overcomplicated solution: pmap

rowmean5 <- beavers %>% map( ~.x%>% 
  mutate(mean_all = pmap_dbl(., function(...) mean(c(...)))))

#  If we want to make use of more than 2 variables (or lists), we can use the pmap function. We need to use ... rather than specifying the columns/lists separately. In the end, we combine all the variables with c()

rowmean5 %>% map(~head(.x))



```


## For these tasks choose your favorite method

**Task 4** Caluclate the mean temp hourwise (900-950,1000-1050 etc.)
```{r}

# get hour
hourwise_mean_temp <- beaver1 %>% mutate(hour = substring(time,1,1))
head(hourwise_mean_temp) 

# group by hour and clac mean
hourwise_mean_temp <- beaver1 %>% mutate(hour = substring(time,1,1)) %>%
  group_by(hour)%>%
  summarise(hour_temp <- mean(temp), .groups = "keep")
head(hourwise_mean_temp) 


# with map loop
hourwise_mean_temp <- beavers %>% map(~.x %>% mutate(hour = substring(time,1,1)) %>%
  group_by(hour)%>%
  summarise(hour_temp = mean(temp), .groups = "keep"))
hourwise_mean_temp %>% map(~head(.x))


# with lapply
hourwise_mean_temp2 <- lapply(beavers, function(x) x %>% mutate(hour = substring(time,1,1)) %>%
  group_by(hour)%>%
  summarise(hour_temp = mean(temp), .groups = "keep") )

hourwise_mean_temp2 %>% map(~head(.x))

```

**Task 5** Make a new list that contains the beaver data of both beavers only for even days
```{r}
even_beavers <- beavers %>% map(~.x[.x$day %% 2 == 0,])

beavers %>% map(~table(.x$day))
even_beavers %>% map(~table(.x$day))

```

**Task 6** Look at the starwars data set column films. This is a list in da data frame! Lets say you want to look at the distribution of skin color and haircolor for the different films. Can you make a list containing one element per film with all characters? How are the distributions filmwise? Make some nice plots (maybe in a loop).
```{r}
library(ggplot2)
head(dplyr::starwars)

# "cheating" with nice tidyverse functions
head(starwars$films)
starwars_by_films <- starwars %>% tidyr::unnest(films) %>% split(., .$films)


map(starwars_by_films, ~ggplot(.x, aes(x=skin_color))+
      geom_bar())

map(starwars_by_films, ~ggplot(.x, aes(x=hair_color))+
      geom_bar())


```

## Benchmark the different options

**Task 7** There is a function microbenchmark from the package microbenchmark, which allows you to compare the efficency of different functions.
Take one of the tasks before and benchmark the three loop types. Which was the best solution?


```{r}
################
# Microbenchmark Task 3
#################

microbenchmark::microbenchmark(for(i in 1:length(beavers))  {
      rowmean[[i]] <- list()
  for(j in 1:nrow(beavers[[i]])){
  rowmean[[i]][[j]] <- mean(as.numeric(beavers[[i]][j,]))
}},

for(i in 1:length(beavers)) {rowmean2[[i]] <- rowMeans(beavers[[i]])},

lapply(beavers, apply, 1, mean), 

beavers %>% map(~mutate(.x, means = rowMeans(.))),

beavers %>% map( ~.x%>% 
  mutate(mean_all = pmap_dbl(., function(...) mean(c(...)))))

                             )



####################
# Microbenchmark Task 4
#####################

microbenchmark::microbenchmark(
beavers %>% map(~.x %>% mutate(hour = substring(time,1,1)) %>%
  group_by(hour)%>%
  summarise(hour_temp = mean(temp), .groups = "keep")),

lapply(beavers, function(x) x %>% mutate(hour = substring(time,1,1)) %>%
  group_by(hour)%>%
  summarise(hour_temp = mean(temp), .groups = "keep") ))







```





