---
title: 'Coding tasks: melt & facet_wrap for ggplot'
author: "Melina Klostermann"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document
---
```{r setup, include=FALSE}
require("knitr")
#xx opts_knit$set(root.dir = "/Users/melinaklostermann/Documents/iCLIP-course")
opts_knit$set(root.dir = "/Users/melinaklostermann/Documents/iCLIP-course")
knitr::opts_chunk$set(warning=FALSE, message=FALSE, cache=TRUE, eval = T, echo = T) #, fig.pos = "!H", out.extra = ""
```

This is the coding task for 29.10. Please do not work on it more then half a day and leave all left over exercises. On the second page I wrote some tips, but you are also welcome to solve the tasks differently.

## Example Data Set
```{r}
library(datasets)
library(ggplot2)
library(tidyverse)

data("airquality")
head(airquality)

```

## 1. Plot the wind on each day in may (month 5)

```{r}
# histogram
p1 <- ggplot(airquality[airquality$Month==5,], aes(x=Day, y=Wind))+
  geom_col()

p1

# dotplot
p2 <- ggplot(airquality[airquality$Month==5,], aes(x=Day, y=Wind))+
  geom_point()

p2

# line
p3 <- ggplot(airquality[airquality$Month==5,], aes(x=Day, y=Wind))+
  geom_line()

p3

```
## 2. Color wind intensities above 10 blue
```{r}
# mark strong wind with dplyr
airquality <- airquality %>% mutate(strong_wind = case_when(
  Wind > 10 ~ T, 
  T~F))

# mark strong wind with base R
airquality$also_strong_wind <- F
airquality[airquality$Wind > 10,]$also_strong_wind <- T


# color in plot
# with color name
p1 <- ggplot(airquality[airquality$Month==5,], aes(x=Day, y=Wind, fill = strong_wind))+
  geom_col()+
  scale_fill_manual(values =c("grey", "blue"))
p1

# with Hex code
p1 <- ggplot(airquality[airquality$Month==5,], aes(x=Day, y=Wind, fill = strong_wind))+
  geom_col()+
  scale_fill_manual(values =c("grey", "#008080"))
p1



```

## 3. Change the theme of the plot
```{r}
# there are many themes to choose from:
p1 + theme_bw()
p1 + theme_classic()
# more themes see: https://www.datanovia.com/en/blog/ggplot-themes-gallery/  

# make your own theme:
p4 <- ggplot(airquality[airquality$Month==5,], aes(x=Day, y=Wind, fill = strong_wind))+
  geom_col()+ 
  scale_fill_manual(values =c("yellow", "blue"))+
  theme(
  # get rid of panel grids
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  # Change plot and panel background
  plot.background=element_rect(fill = "gray"),
  panel.background = element_rect(fill = 'black'),
  # Change legend 
  legend.position = c(0.6, 0.07),
  legend.direction = "horizontal",
  legend.background = element_rect(fill = "black", color = NA),
  legend.key = element_rect(color = "gray", fill = "black"),
  legend.title = element_text(color = "white"),
  legend.text = element_text(color = "white")
  )

p4  


```
## 4. Use facet_wrap to plot the wind in each month
```{r}
p1 <- ggplot(airquality, aes(x=Day, y=Wind, fill = strong_wind))+
  geom_col()+
  scale_fill_manual(values =c("grey", "blue"))+
  facet_wrap(~Month)+ # tell facet_wrap which column contains the facets
  theme_bw()

p1


```

## 5. Plot ozone, solar.r wind and temp for month 5 as facets of the same plot

```{r}
# you can use melt to get all values for ozone, solar.r wind and temp in one column

airquality_melt <- reshape2::melt(airquality, id.vars=c("Day", "Month", "strong_wind", "also_strong_wind"), value.name = "messured", variable.name="air_factor")

p5 <- ggplot(airquality_melt[airquality_melt$Month==5,], aes(x=Day, y=messured, fill = air_factor))+
  geom_col()+
  facet_wrap(~air_factor, scales = "free_y")+
  scale_fill_manual(values = c("darkblue", "yellow", "#0000FF", "red"))+
  theme_dark()

p5


```
## 6. Make a plot with subplot for each condition and each month
With facet_grid you can split up the plot by two columns.

```{r}
p5 <- ggplot(airquality_melt, aes(x=Day, y=messured, fill = air_factor))+
  geom_col()+
  facet_grid(Month~air_factor, scales = "free")+
  scale_fill_manual(values = c("darkblue", "yellow", "#0000FF", "red"))+
  theme_dark()

p5

```

## 7. Plot the temperatures of each month and color by the temperature (warmest red, coldest blue)

```{r, fig.width=10, fig.height=3}
p7 <- ggplot(airquality, aes(x=Day, y=Temp, fill = Temp, color = Temp))+
  geom_col()+
  facet_wrap(~Month, ncol = 5)+
  scale_fill_gradient(low="blue", high="red")+
  scale_color_gradient(low="blue", high="red")+
  theme_minimal()
p7


```


\newpage

# Tips
## 1.
You can use the filter function from dplyr

## 2.
Make a new column high_wind and use color = high_wind in the aes() of your ggplot

## 3.
See https://ggplot2.tidyverse.org/reference/ggtheme.html

## 4.
With facet wrap you can split up one plot into multiple by one variable. The syntax is </p>
```{r eval = F}
your_ggplot+ 
facet_wrap(~colname_to_split)
```

## 5.
You can use melt from the reshape2 package. It will melt all columns together. Try it out. </p>
Now for the plot you want to keep month and day unmelted. You can specify id.vars which will not be melted. You can use value as y day as x and variable for facet_wrap.

## 6.
```{r eval = F}
your_ggplot+ 
facet_grid(col1~col2)
```

## 7.
```{r eval = F}
your_ggplot+ 
scale_colour_gradient()  or scale_fill_gradient()
```



