---
title: "Some practices of dplyr 20.09.2021"
author: "You Zhou"
date: "20/09/2021"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: no
    toc: yes
    toc_depth: 3
  bookdown::html_document2:
    fig_caption: yes
    number_sections: no
    theme: cerulean
    toc: yes
    toc_depth: 3
graphics: yes
header-includes:
- \makeatletter\renewcommand*{\fps@figure}{h}\makeatother
- \usepackage{placeins}
geometry: margin=1in
fontsize: 18pt
---

```{r knitr off, echo=FALSE, message=FALSE,cache=FALSE, results="hide", render=FALSE ,warning=FALSE}
knitr::opts_chunk$set(echo=TRUE, error=FALSE, warning=FALSE,message=FALSE)
# Setting the global options for all chunks
```

\FloatBarrier

# Introduction
In this coding tasks, we will do some practices about using the `dplyr` package. Before moving to 
the questions, please make sure that you load the library and use the correct test data set ; )

```{r, echo=T}
## loading the required package
library(tidyverse) ## this library is a libraries collection which includes the 'dplyr'
## loading the correct data set for the incoming questions
data("diamonds")
head(diamonds)
```

# Practice 1
Extract the carat, cut, color and price column from the data.

Tips: `select` function may help.

```{r}
df <- diamonds %>% select(carat:color, price)
df
```

# Practice 2
Making a new column to show the number of row for all the diamonds with function `mutate`.

```{r}
df <- df %>% mutate(row_number = row_number())
df
```

# Practice 3
Let's first make a unique ID for each diamond.

```{r echo=TRUE}
set.seed(1)
diamonds$ID <- sample(1:nrow(diamonds), nrow(diamonds), replace = FALSE)
```

And we have a new data which have the column ID, virtual price and some number.

```{r echo=TRUE}
df_virtual <- data.frame(
  ID = 1:nrow(diamonds), 
  virtual_price = nrow(diamonds):1,
  some_number = nrow(diamonds):1
)
head(df_virtual)
```

Now the question is how can we joint the `df_virtual` to the original `diamonds` data frame.

```{r}
df <- left_join(diamonds, df_virtual, by = "ID") %>% 
  select(!(x:z))
head(df)
```

# Practice 4
Find out the maximum `price` for different `color` and convert the result into a bar plot.

Tips: `group_by` function may help.

```{r}
## this package is unnecessary but can make the plot nicer ; )
library(ggthemr)
ggthemr('solarized')

df <- diamonds %>% group_by(color) %>% 
  summarise(price = max(price))

ggplot(df, aes(x = color, y = price, label = price)) + 
  geom_bar(stat = "identity") + 
  geom_text()
```

# Practice 5
Sum up all the price for the diamonds with the same `cut` and `color`. And using a `dodge bar` 
chart for the visualization.

```{r}
df <- diamonds %>% group_by(cut, color) %>%
  summarise(Price_sum = sum(price)) 

ggplot(df, aes(x = color, y = Price_sum, fill = cut)) + 
  geom_bar(stat = "identity", position = "dodge")
```

# Practice 6
For the diamonds with the **Premium** `cut`, extract the **top 5** diamonds with the highest 
`price` for each `color`. Then make a *boxplot* to show this result.

```{r}
df <- diamonds %>% arrange(desc(price)) %>% 
  filter(cut == "Premium") %>% 
  group_by(color) %>% 
  slice_max(order_by = price, n = 5)
  
ggplot(df, aes(x = color, y = price)) + 
  geom_boxplot()
```

# Practice 7
Count the number for the diamonds which have the same `cut` and `color`.

```{r}
df <- diamonds %>% group_by(cut, color) %>%
  summarise(Number = n())
head(df)
```
